Penalized regression for left-truncated and right-censored survival data

### 2021-04-16 manuscript.tex

\author[1]{Sarah F. McGough\thanks{Contributed equally.}}
\author[1]{Devin Incerti\samethanks}
\author[1]{Svetlana Lyalina}
\author[1]{Ryan Copping}
\author[2, 3]{Balasubramanian Narasimhan}
\author[2, 3]{Robert Tibshirani}
\affil[1]{Genentech, Inc, South San Francisco, CA, USA}
\affil[2]{Department of Statistics, Stanford University, Stanford, CA, USA}
\affil[3]{Department of Biomedical Data Sciences, Stanford University, Stanford, CA, USA}

@@

Simon et. al\cite{simon2011regularization} introduced a fast, pathwise algorithm to fit a regularized Cox proportional hazards model solving for

@@

where $\beta$ is the vector of regression coefficients and is maximized over \autoref{eqn:simon} subject to a constraint defined by the penalty $\lambda \cdot P_{\alpha}\beta$. As described elsewhere \cite{simon2011regularization}, the penalty can take the form of the $l_1$ (lasso), $l_2$ (ridge), or elastic net penalty,  and the algorithm fits over a path of $\lambda$ values via cyclical coordinate descent. Instead of solving a least squares problem at each iteration, the algorithm solves a penalized reweighted least squares problem, in which observation weights $w$ are defined in the coordinate descent least squares minimization step. The optimal $\lambda$ can be identified through cross-validation. Tied survival times are handled via the Breslow approximation \cite{breslow1972} and details provided in \cite{simon2011regularization}.

@@

\autoref{fig:hr-plot} displays hazard ratios for each model. The plot for the large $p$ model only contains coefficients ranked in the top 10 by absolute value of the hazard ratio in either the left truncation adjusted or non-adjusted model.  The arrows denote the impact of adjusting for left truncation on the hazard ratio. In some cases the adjustment can have a considerable impact on the hazard ratios, often moving a large hazard ratio all the way to $1$ (i.e., to a coefficient of $0$). This is most notable with diagnosis year, which has a hazard ratio of approximately $1.5$ in both the small and large $p$ models but is set to $1$ after the left truncation adjustment. This effect is likely due to the strong negative correlation between calendar time and the probability that a patient is left-truncated.

@@

\autoref{tbl:model-discrimination}  and \autoref{fig:surv-calibration} evaluate the models based on predictions on the test set. \autoref{tbl:model-discrimination} compares models using the C-index. The C-index is relatively low in all of these illustrative models, though is artificially higher in the unadjusted models that do not account for left truncation---sometimes considerably so---and decreases after adjustment as the predictor hazard ratios are pushed toward $1$ and there are fewer differences in predicted relative risks across patients. As expected, the lasso model has a higher C-index in the larger $p$ setting as the Cox model is considerably overfitted.

@@

In our real-world use case, an analyst could have mistakenly concluded that the high C-index in the non-adjusted model was indicative of good model performance when it was in fact driven primarily by a high correlation between predictors and left truncation. For instance, in our large $p$ model the hazard ratio for the predictor "year of diagnosis" moved from being the largest in absolute value to $1$ after adjusting for left truncation. The high C-index in the non-adjusted model was artificial because patients diagnosed earlier appeared to survive longer due to immortal time bias. To facilitate model interpretation, we recommend that models should be evaluated using both discrimination and calibration based metrics.

>>>>

\author[1]{Sarah F. McGough\thanks{Contributed equally.}}
\author[1]{Devin Incerti\samethanks}
\author[1]{Svetlana Lyalina}
\author[1]{Ryan Copping}
\author[2, 3]{Balasubramanian Narasimhan}
\author[2, 3]{Robert Tibshirani}
\affil[1]{Genentech, Inc, South San Francisco, CA, USA}
\affil[2]{Department of Statistics, Stanford University, Stanford, CA, USA}
\affil[3]{Department of Biomedical Data Science, Stanford University, Stanford, CA, USA}

@@

Simon \emph{et al.} \cite{simon2011regularization} introduced a fast, pathwise algorithm to fit a regularized Cox proportional hazards model solving for

@@

where $\beta$ is the vector of regression coefficients and is maximized over \autoref{eqn:simon} subject to a constraint defined by $\lambda \cdot P_{\alpha}\beta$, and $P_{\alpha}\beta$ is the $l_1$ (lasso), $l_2$ (ridge), or elastic net penalty on the sum of coefficients $\beta$ described elsewhere \cite{simon2011regularization, tibshirani1996regression, tikhonov1963ridge, zou2005regularization}. The algorithm by Simon \emph{et al.} fits over a path of $\lambda$ values via cyclical coordinate descent. Instead of solving a least squares problem at each iteration, the algorithm solves a penalized reweighted least squares problem, in which observation weights $w$ are defined in the coordinate descent least squares minimization step. The optimal $\lambda$ can be identified through cross-validation. Tied survival times are handled via the Breslow approximation \cite{breslow1972} and details provided in \cite{simon2011regularization}.

@@

\autoref{fig:hr-plot} displays hazard ratios for each model. The plot for the large $p$ model only contains coefficients ranked in the top 10 by absolute value of the hazard ratio in either the left truncation adjusted or non-adjusted model.  The arrows denote the impact of adjusting for left truncation on the hazard ratio. In some cases the adjustment can have a considerable impact on the hazard ratios, often moving a large hazard ratio all the way to $1$ (i.e., to a coefficient of $0$). This is most notable with diagnosis year, which has a hazard ratio of approximately $1.5$ in both the small and large $p$ models but is reduced to $1$ or less than $1$ after the left truncation adjustment. This effect is likely due to the strong negative correlation between calendar time and the probability that a patient is left-truncated.

@@

\autoref{tbl:model-discrimination}  and \autoref{fig:surv-calibration} evaluate the models based on predictions on the test set. \autoref{tbl:model-discrimination} compares models using the C-index. The C-index is relatively low in all of these illustrative models, though is artificially higher in the unadjusted models that do not account for left truncation---sometimes considerably so---and decreases after adjustment as there are fewer differences in predicted relative risks across patients and the discriminatory ability of the model is low. As expected, the lasso model has a higher C-index in the larger $p$ setting as the Cox model is considerably overfitted.

@@

In our real-world use case, an analyst could have mistakenly concluded that the high C-index in the non-adjusted model was indicative of good model performance when it was in fact driven primarily by a high correlation between predictors and left truncation. For instance, in our large $p$ model the hazard ratio for the predictor "year of diagnosis" moved from being the largest in absolute value to around $1$ after adjusting for left truncation. The strongly positive effect of "year of diagnosis" on the hazard in the non-adjusted model was artificial because patients diagnosed earlier appeared to survive longer due to immortal time bias. This effect is reversed completely after the left truncation adjustment, and illustrates the bias to coefficient estimates that is driven by the association between predictors and left truncation time. In the case of "year of diagnosis," this association is negative and drives an upward bias to the hazard ratio. While many of the other hazard ratios in the large $p$ model were pushed towards $1$ after left truncation adjustment, the direction of this bias depends on the direction of the association with left truncation time and varies by predictor. Because the high C-index misleadingly suggested good model performance, we recommend that models should be evaluated using both discrimination and calibration based metrics to facilitate interpretation.
